The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-07-14 IEEE SP Magazine/07-2021 - Draft/article.tex
Finished the computational example section.

We propose to measure the discriminative power of the CQT-SECs by evaluating them on the NSynth dataset, a large-scale and high-quality dataset of annotated musical notes which is publicly available\footnote{https://magenta.tensorflow.org/datasets/nsynth} \cite{engel2017}. The NSynth dataset is composed of 305,979 musical notes which were generated from 1,006 instruments as 4-second, monophonic audio signals at a sampling rate of 16 kHz, with pitches ranging over all the note numbers of a standard MIDI piano (21-108) and with 5 different velocities (25, 50, 75, 100, and 127), whenever applicable. The instruments are organized into 11 families (bass, brass, flute, guitar, keyboard, mallet, organ, reed, string, synth\_lead, and vocal) and 3 sources (acoustic, electronic, and synthetic). We chose to evaluate the CQT-SECs on the notes with a velocity of 75 only, leading to 60,388 different notes for 945 different instruments.

We derived the CQT-SECs for all the notes in this subset, by first computing the CQT spectrogram using librosa, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave, and then extracting 20 coefficients from the envelope resulting from the deconvolution of the CQT spectrogram, leading to CQT-SECs of size 20 by 126 (time frames). We used a power CQT spectrogram (i.e., magnitude to the power of 2) and took the real part of the envelope to ensure real values. For comparison, we also computed the MFCCs, using librosa, with a sampling rate of 16 kHz, a window length of 1024 samples, a hop length of 512 samples, and 20 coefficients, leading to MFCCs of size 20 by 126, just like with the CQT-SECs.

We computed the cosine similarity for every pair of CQT-SECs and every pair of MFCCs (without repetition), after flattening the features into vectors of length 2520 (20 times 126). We then averaged these note similarities over every one of the 945 different instruments, leading to similarity matrices of size 945 by 945, for both the CQT-SECs and the MFCCs. Figure \ref{fig:instrument_similarities} shows these instrument similarity matrices for the CQT-SECs and the MFCCs. As we can see, the similarities for the CQT-SECs have more variance, while the similarities for the MFCCs are mostly very high (close to 1) showing poor discriminative power between different instruments. Figure \ref{fig:instrument_similarities2} additionally shows the self-similarities, i.e., the diagonal in the similarity matrix (in green) and the error bars for the cross-similarities, i.e., the off-diagonal values in the similarity matrix (means in red and standard deviations in yellow) for every instrument, for the CQT-SECs and the MFCCs. As we can see, the self-similarities for the CQT-SECs are noticeably higher that the mean of the cross-similarities for most of the instruments, and generally higher that the errors as well, showing good discriminative power, while the self-similarities and cross-similarities for the MFCCs are all very high (close to 1).

@@

We also averaged the note similarities over every one of the 11 instrument families, leading to similarity matrices of size 11 by 11, for both the CQT-SECs and the MFCCs. Figure \ref{fig:family_similarities} shows these instrument family similarity matrices for the CQT-SECs and the MFCCs. As we can see here as well, the similarities for the CQT-SECs have more variance, ... while the similarities for the MFCCs are mostly very high. Figure \ref{fig:family_similarities2} additionally shows the self-similarities (in green) and the error bars for the cross-similarities (means and standard deviations in red) for every instrument family, for the CQT-SECs and the MFCCs. As we can see, the self-similarities for the CQT-SECs are noticeably higher that the mean of the cross-similarities for most of the instrument families, and generally higher that the errors as well, while the self-similarities and cross-similarities for the MFCCs are all very high.

>>>>

We propose to measure the discriminative power of the CQT-SECs by evaluating them for a simple instrument similarity task on the NSynth dataset \cite{engel2017}, a large-scale and high-quality dataset of annotated musical notes which is publicly available\footnote{https://magenta.tensorflow.org/datasets/nsynth}. The NSynth dataset is composed of 305,979 musical notes which were generated from 1,006 instruments as 4-second, monophonic audio signals at a sampling rate of 16 kHz, with pitches ranging over all the note numbers of a standard MIDI piano (21-108) and with 5 different velocities (25, 50, 75, 100, and 127), whenever applicable. The instruments are organized into 11 families, namely, bass, brass, flute, guitar, keyboard, mallet, organ, reed, string, synth\_lead, and vocal, and 3 sources, namely, acoustic, electronic, and synthetic. We chose to evaluate the CQT-SECs on the notes with a velocity of 75 only, leading to a subset of 60,388 different notes for 945 different instruments.

We derived the CQT-SECs for all the notes in this subset, by first computing the CQT spectrogram using librosa, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave, and then extracting 20 coefficients from the envelope resulting from the deconvolution of the CQT spectrogram, leading to CQT-SECs of size 20 by 126 (time frames). We used a power CQT spectrogram (i.e., magnitude to the power of 2) and took the real part of the envelope to ensure real values. For comparison, we also computed the MFCCs, using librosa, with a sampling rate of 16 kHz, a window length of 1024 samples, a hop length of 512 samples, and 20 coefficients, leading to MFCCs of same size as the CQT-SECs, i.e., 20 by 126.

We then computed the cosine similarity for every pair of CQT-SECs and every pair of MFCCs (without repetition), after flattening the features into vectors of length 2520 (20 times 126). We averaged these note similarities over every one of the 945 instruments, leading to similarity matrices of size 945 by 945, for both the CQT-SECs and the MFCCs. Figure \ref{fig:instrument_similarities} shows these instrument similarity matrices for the CQT-SECs and the MFCCs. As we can see, the similarities for the CQT-SECs have more variance, while the similarities for the MFCCs are mostly very high (close to 1) showing poor discriminative power between the different instruments. Figure \ref{fig:instrument_similarities2} additionally shows the self-similarities, i.e., the diagonal values in the similarity matrix (in green) and the error bars for the cross-similarities, i.e., the off-diagonal values in the similarity matrix (means in red and standard deviations in yellow) for every instrument, for the CQT-SECs and the MFCCs. As we can see, the self-similarities for the CQT-SECs are noticeably higher than the means of the cross-similarities for most of the instruments, and generally higher that the means plus standard deviations as well, showing good discriminative power, while the self-similarities and cross-similarities for the MFCCs are all very high (close to 1).

@@

We also averaged the note similarities over every one of the 11 instrument families, leading to similarity matrices of size 11 by 11, for both the CQT-SECs and the MFCCs. Figure \ref{fig:family_similarities} shows these instrument family similarity matrices for the CQT-SECs and the MFCCs. As we can see here as well, the similarities for the CQT-SECs have more variance, showing mostly clearer high self-similarities and low cross-similarities, while the similarities for the MFCCs are mostly very high. Figure \ref{fig:family_similarities2} additionally shows the self-similarities (in green) and the error bars for the cross-similarities (means and standard deviations in red) for every instrument family, for the CQT-SECs and the MFCCs. As we can again see, the self-similarities for the CQT-SECs are noticeably and mostly higher than the means plus standard deviations (except for the string and synth\_lead families), while the self-similarities and cross-similarities for the MFCCs are all very high and harder to tell apart.
